\clearpage
\addcontentsline{toc}{chapter}{Introduction}
\chapter*{Introduction}

The \nutils{} are a set of \unix{} shell programs for working with phylogenetic trees. Their main features are:
\begin{itemize}
 \item they require no user interaction\footnote{Why this is a good thing is not the focus of this document: I shall assume that if you are reading this, you already know when a command-line interface is better than an interactive interface.}
 \item they can work on any number of trees at a time
 \item they perform reasonably well with large trees
 \item they are implemented as filters
\end{itemize}
They are not tools for \emph{making} phylogenies. Rather, they are for
processing existing ones, by which I mean manipulating the tree or extracting
information from it: rerooting, simplifying, extracting subtrees, printing
branch lengths and distances, etc - see table \ref{tbl_prog_list}; a glance
at the table of contents should also give you an idea.

Each of the programs performs one task (with some variants). For example, here
is how you would reroot a series of phylograms contained in file
\texttt{mytrees.nw} using node \texttt{Dmelano} as the outgroup:

\begin{verbatim}
$ nw_reroot mytrees.nw Dmelano
\end{verbatim} 
Now, you might want to make cladograms from the rerooted trees. Program \topology{} does the job, and since the utilities are filters, you can do it in a single command:
\begin{verbatim}
$ nw_reroot mytrees.nw Dmelano | nw_topology -
\end{verbatim}
As you can see, it is straightforward to pipe \nutils{} together, and of course they can be mixed freely with any other shell program (see e.g. \ref{sct_counting_leaves}).

\subsection*{Organization of This Document}

This document is organized as follows: chapter \ref{chap_general} discusses
common features of the \nutils, chapter \ref{chap_simple} shows examples of
simple tasks, and chapter \ref{chap_adv} has examples of more advanced tasks. 

It is not necessary to read this material in order: you can pretty much jump to
any section in chapters \ref{chap_simple} and \ref{chap_adv}, they do not
require reading previous sections. I would suggest reading chapter
\ref{chap_general}, then section \ref{sct_display} because it explains how all
the tree graphics were produced.

The \nw{} files for all the examples in this tutorial can be found in
subdirectory \texttt{data}.

\begin{table}[b]
\begin{tabular}{ll}
{\bf Program} & {\bf Function } \\
\hline
\clade	&	Extracts subtrees defined by labels\\
\condense	&	Condenses (simplifies) trees \\
\display	&	Shows trees as graphs (\ascii{} graphics or \svg) \\
\distance	&	Prints distances between nodes, in various ways \\
\ed	&	Stream editor (\textit{\`{a} la} \texttt{sed} or \texttt{awk}) \\
\gen	&	Random tree generator \\
\nwindent	&	Shows Newick in indented form \\ 
\labels	&	Prints node labels \\
\match	&	Finds matches of a tree in another one \\
\order	&	Orders tree (preserving topology) \\
\prune	&	Removes branches based on labels \\ 
\reroot	&	(Re)roots the tree \\
\support	&	Computes bootstrap support of a tree given replicate trees \\
\topology & Alters branch properties, preserving topology \\
\trim & Trims a tree at a specified depth
\end{tabular}	
\caption{The \nutils{} and their functions}
\label{tbl_prog_list}
\end{table}
