\lstset{
	language=Python,
	basicstyle=\ttfamily,
	keywordstyle=\color{NavyBlue},
	stringstyle=\color{OliveGreen},
	commentstyle=\color{red},
	frame=single,
	frameround=tttt,
	framexleftmargin=8mm,
	numbers=left
}

\chapter{Python Bindings}
\label{chap_python_lib}

Although the \nutils{} are primarily designed for shell use, it is also possible
to use their functions from Python programs: all the core functionality of the
utilities is bundled in a C library, \texttt{libnw}, which can be accessed
through Python's \texttt{ctypes} module. The distribution contains a file,
\texttt{newick\_utils.py}, that provides the Python to C mappings; it also
builds an object-oriented interface over it.

Let's say we want to add a utility that prints simple statistics about trees,
like the number of nodes, the depth, whether it is a cladogram or a phylogram,
etc (in other words, a Python version of \stats). We will call it
\texttt{nw\_info.py}, and we'll pass it a \nw{} file on standard input, so the
usage will be something like:

\begin{verbatim}
$ nw_info.py < data/catarrhini
\end{verbatim}

\noindent{}The overall structure of this program is simple: iteratively read
all the input trees, and do something with each of them:

\begin{lstlisting}
from newick_utils import *

for tree in Tree.parse_newick_input():
    pass # process tree here!
\end{lstlisting}

\noindent{}Line 1 imports definitions from the \texttt{newick\_utils.py}
module. Line 3 is the main loop: the \texttt{Tree.parse\_newick\_input}
reads standard input and yields an instance of class \texttt{Tree} for each
Newick string. We can now work with it, using methods of class \texttt{Tree} or adding our own:

\lstinputlisting{../src/nw_info.py}

\noindent{}When we run the program, we get:

\verbatiminput{python_1_txt.cmd}
\verbatiminput{python_1_txt.out}

As you can see, most of the work is done by methods called on the \texttt{tree}
object, such as \texttt{get\_leaf\_count} which (surprise!) returns the number
of leaves of a tree. But since there is no method for counting polytomies, we
added our own function, \texttt{count\_polytomies}, which takes a \texttt{Tree}
object as argument.

As another example, a simple implementation of \reroot{} is found in
\texttt{src/nw\_reroot.py}. It demonstrates two approaches: a heavily
object-oriented one, in which the user mainly calls methods on Python objects,
and a "thin" one, in which the calls are essentially to C functions through
\texttt{libnw}. While not as fast as \reroot{}, its performance is still quite
acceptable, especially in "thin" mode.

\section{API Documentation}

\noindent{}Detailed information about all classes and methods available for accessing the \nutils{} library from Python is found in file \texttt{newick\_utils.py}. Note that the library must be installed on your system, which means that you must compile from source.
