\documentclass[a4paper,10pt]{report}

\usepackage{url}

% Title Page
\title{Newick Utilities Tutorial}
\author{Thomas Junier}

\newcommand{\nutils}{Newick Utilities}
\newcommand{\unix}{\textsc{Unix}}
\newcommand{\ascii}{\textsc{ASCII}}
\newcommand{\svg}{\textsc{SVG}}
\newcommand{\display}{\texttt{nw\_display}}
\newcommand{\stdin}{standard input}

\begin{document}
\maketitle
\tableofcontents

\chapter{Introduction}

The \nutils{} are a set of \unix{} shell programs for working with Newick-formatted phylogenetic trees. Their main feature is that they require no user interaction. Why this is a good thing is not the focus of this document: I shall assume that if you are reading this, you already know when a command-line interface is better than an interactive interface.

The \nutils{} are not tools for \emph{making} phylogenies. Rather, they are for working with existing ones, by which I mean manipulating the tree or extracting information from it: rerooting, simplifying, extracting subtrees, printing branch lengths and distances, etc - a glance at the table of contents of this document shoul dgive you an idea.

Each of the program performs one task (with some variants). For example, here is how you would reroot a series of phylograms contained in file \texttt{mytrees.nw} using node \texttt{Dmelano} as outgroup:

\begin{verbatim}
$ nw_reroot mytrees.nw Dmelano
\end{verbatim} 
Now, you might want to make cladograms from the rerooted trees. This would be done like this:
\begin{verbatim}
$ nw_reroot mytrees.nw Dmelano | nw_topology -
\end{verbatim}

This document is organized as follows: chapter \ref{chap_general} discusses common features of the \nutils, chapter \ref{chap_simple} shows examples of simple tasks, and chapter \ref{chap_adv} has examples of more advanced tasks.

\section{Conventions}

Most of the utilities produce Newick output. This is a rather terse format, not easily readable for humans. Therefore, I will generally show the result in graph form, either as \ascii{} or as (rendered!) \svg. The conversion from Newick to graph will be performed with \display{}. To avoid tedious repetitions, however, I will not always explicitly show the \display{} step; that is, if I say that 
\begin{verbatim}
$ nw_clade catarrhini.nw Pongo Pan
\end{verbatim} 
results in this:
\begin{samepage}
\begin{verbatim}
                  +-----------------+ Gorilla     
                  |                               
 +----------------+ Homininae +----------+ Pan    
 |                +-----------+ Hominini          
=| Hominidae                  +----------+ Homo   
 |                                                
 +---------------------------------+ Pongo  
\end{verbatim} 
\end{samepage}
it will mean that the last step was \display{}, something like this:
\begin{verbatim}
$ nw_clade catarrhini.nw Pongo Pan | nw_display -
\end{verbatim}
If the graph is \svg, then it will have been \display{} \verb+-s+.

\chapter{Generalities}
\label{chap_general}

\section{Help}
\label{sect_help}

All programs in the \nutils{} package will print a help message if passed option \texttt{-h}:

\begin{samepage}
\begin{verbatim}
$ nw_indent -h
Indents the Newick, making structure more clear.

Synopsis
--------

nw_indent [-cht:] <newick trees filename|->

Input
-----

Argument is the name of a file that contains Newick trees, or '-' (in
which case trees are read from standard input).

Output
------

By default, prints the input tree, with each parenthesis and each leaf on a
line of its own, and indented a multiple of '  ' (two spaces) to reflect
structure. The default output is valid Newick.
[...]
\end{verbatim}
\end{samepage}
The help page describes the program's purpose, its input and output, and its options, in a format reminiscent of \unix{} manpages. It also shows a few examples. All examples can be tried out using files in the \texttt{data} directory.

\section{Input}
\label{sect_input}

Since the \nutils{} are for working with trees, it should not be a surprise that all programs take Newick as input. The Newick format is a file format for (phylogenetic) trees. It is one of the most widely used and uderstood formats.
A complete description can be found at \url{http://evolution.genetics.washington.edu/phylip/newicktree.html}.

The input tree(s) are always the first argument to the program (after any options). They may either be stored in a file, or piped on \stdin{}. In the latter case, the filename is replaced by a '\texttt{-}' (dash):

\begin{samepage}
\begin{verbatim}
$ nw_display mytrees.nw
\end{verbatim}
is the same as
\begin{verbatim}
$ cat mytrees.nw | nw_display -
\end{verbatim}
\end{samepage}
Of course the second form is only really useful when chaining several programs into pipelines.

Most programs can work on any number of trees. That is, in the example above the file \texttt{mytrees.nw} may contain one or more trees, preferably one tree per line\footnote{It may also work if the trees are formatted differently, the programs were only tested on input files with one per line.}. The task will be performed on all input trees. So if you need to reroot 1,000 trees on the same outgroup, you can do it all in a single step (see \ref{sct_reroot}).

\section{Output}
\label{sect_output}

The \nutils{} either modify trees and print the result, or print information about the trees. In the first case, the output is also Newick. In the 

\chapter{Simple Tasks}
\label{chap_simple}

\section{Rerooting}
\label{sct_reroot}

\section{Labels}

\chapter{Advanced Tasks}
\label{chap_adv}
\end{document}
